---
title: Principles of Statistical Data Analysis
author:
- Jiacheng Shen
- Sajjad Ayashm
date: "2024-10-28"
output:
  bookdown::pdf_document2:
    toc: false
header-includes:
  - \usepackage{float}
subtitle: Home Work 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
```
```{r global-options, include=FALSE}
options(digits = 3) 
```

# Data Exploration
In this section, we explore the data using summary statistics and visualizations to understand the distribution of ant abundance and soil moisture. Table \@ref(tab:ants-data) shows 5 rows of our data.


```{r ants-data, echo=FALSE}
load("ants.RData")
kable(head(ants), caption = 'Data in ants.RData file')  %>%
  kable_styling(latex_options = "H")
```

## Data Summary:
Table \@ref(tab:ants-summary) shows the first few rows of the data set.

```{r ants-summary, eval=TRUE, echo=FALSE}
# Numerical Summary Statistics
abundance_summ <- as.vector(summary(ants$abundance))
moisture_summ <- as.vector(summary(ants$moisture))
# Combine summaries into a single data frame and transpose
combined_summ <- data.frame(
  Statistic = names(summary(ants$abundance)),
  Abundance = abundance_summ,
  Moisture = moisture_summ
)
# Bin moisture levels and calculate mean and variance for each bin
moisture_bins <- cut(ants$moisture, breaks = 5)
mean_abundance <- tapply(ants$abundance, moisture_bins, mean)
variance_abundance <- tapply(ants$abundance, moisture_bins, var)
# Transpose the table for horizontal display
combined_summ_t <- t(combined_summ)
colnames(combined_summ_t) <- combined_summ$Statistic
combined_summ_t <- combined_summ_t[-1, ]
# Display transposed summary table with caption
kable(combined_summ_t, caption = 'Summary of Abundance of Ants and Moisture of Soil') %>%
  kable_styling(latex_options = "H")
```

Additionally, the variance of ant abundance in the sample is **`r format(var(ants$abundance), , scientific = TRUE)`**, and for moisture, it is **`r var(ants$moisture)`**

## Data Visualization
We visualized the distribution of ant abundance using a histogram (Figure \@ref(fig:hist-abundance)) and boxplot (Figure \@ref(fig:boxplot-abundance)) to identify its spread and any potential outliers. A scatter plot of soil moisture versus ant abundance (Figure \@ref(fig:scatter-plot)) was created to explore their relationship. Additionally, a mean-variance plot (Figure \@ref(fig:mean-variance)) was used to assess overdispersion, which supports the choice of a negative binomial model for this data.

```{r hist-abundance, echo=FALSE, fig.cap="Histogram of Ant Abundance", fig.pos="H", fig.width=5, fig.height=4,}
# Histogram of Abundance
hist(ants$abundance, xlab = "Abundance", ylab = "Frequency")
```

```{r boxplot-abundance, echo=FALSE, fig.cap="Boxplot of Abundance", fig.pos="H", fig.width=5, fig.height=4,}
# Boxplot of Abundance
boxplot(ants$abundance, ylab = "Abundance")
```

```{r scatter-plot, echo=FALSE, fig.cap="Scatter Plot of Moisture vs Ant Abundance", fig.pos="H", fig.width=5, fig.height=4,}
# Visualize the Data
plot(ants$moisture, ants$abundance, 
     xlab = "Moisture (%)", ylab = "Ant Abundance")
```

```{r mean-variance, echo=FALSE, fig.cap="Mean vs Variance of Ant Abundance", fig.pos="H", fig.width=5, fig.height=4,}
# Examine Mean-Variance Relationship
mean_abundance <- tapply(ants$abundance, cut(ants$moisture, breaks=5), mean)
variance_abundance <- tapply(ants$abundance, cut(ants$moisture, breaks=5), var)
plot(mean_abundance, variance_abundance, 
     xlab = "Mean Abundance", ylab = "Variance")

```

# Likelihood Function

The likelihood function is shown in Equation.

$$ 
L(\beta_0, \beta_1, \phi) = \prod_{i=1}^n \frac{\Gamma(y_i + 1 / \phi)}{\Gamma(1 / \phi) \, y_i!} \left( \frac{1}{1 + \mu_i \phi} \right)^{1 / \phi} \left( \frac{\mu_i \phi}{1 + \mu_i \phi} \right)^{y_i}
$$ 


## Likelihood Function

To define the likelihood function in R, we use a function that computes the product of probabilities for all observations.
```{r eval=TRUE,echo=FALSE}
# Define the likelihood function
likelihood_function <- function(params, abundance, moisture) {
  beta_0 <- params[1]  # The intercept parameter
  beta_1 <- params[2]  # The coefficient for moisture
  phi <- params[3]     # The overdispersion parameter
  
  # Calculate the mean (mu) for each observation
  mu <- exp(beta_0 + beta_1 * moisture)
  
  # Compute the negative binomial likelihood for each observation
  likelihoods <- dnbinom(abundance, size = 1/phi, mu = mu, log = FALSE)
  
  # Calculate the total likelihood by multiplying individual likelihoods
  total_likelihood <- prod(likelihoods)
  
  return(total_likelihood)
}

initial_params <- c(1, 0.1, 0.5)  # Example starting values for beta_0, beta_1, and phi
likelihood_value <- likelihood_function(initial_params, ants$abundance, ants$moisture)
```


**Note:** If we use the set of parameters described in the example above, the likelihood value results `r likelihood_value`. One possible reason is that the number is too small to exhibit. That's why we derive the following log likelihood function.

## Log-Likelihood Function

The log-likelihood function computes the sum of the log-probabilities instead of multiplying the probabilities directly, providing more numerical stability.

```{r task2 - log, echo=FALSE}
# Define the log-likelihood function
log_likelihood <- function(params, abundance, moisture) {
  beta_0 <- params[1]  # The intercept parameter
  beta_1 <- params[2]  # The coefficient for moisture
  phi <- params[3]     # The overdispersion parameter
  
  # Calculate the mean (mu) for each observation
  mu <- exp(beta_0 + beta_1 * moisture)
  
  # Compute the log-likelihood for each observation
  log_lik <- sum(
    lgamma(abundance + 1 / phi) - lgamma(1 / phi) - lgamma(abundance + 1) +
    (1 / phi) * log(1 / (1 + mu * phi)) +
    abundance * log(mu * phi / (1 + mu * phi))
  )
  
  return(log_lik)  # Return the log-likelihood
}

initial_params <- c(1, 0.1, 0.5)  # Example starting values for beta_0, beta_1, and phi
log_likelihood_value <- log_likelihood(initial_params, ants$abundance, ants$moisture)
```

Using the same set of parameters, the log likelihood values result `r log_likelihood_value`, which makes more sense.

**Explanation:**
Likelihood vs. Log-Likelihood: The likelihood function directly computes the product of individual probabilities, which can lead to underflow when dealing with very small values. The log-likelihood function sums the logarithms of the probabilities, providing greater numerical stability.


# Estimating Equation for \(\beta_1\)

Our objective is to derive the estimating equation for \(\beta_1\) by maximizing the log-likelihood function. The maximum likelihood estimate (MLE) for \(\beta_1\) can be obtained by setting the derivative of the log-likelihood with respect to  \(\beta_1\) to zero, which requires numerical optimization.


The log-likelihood function is given by:

$$
\ell(\beta_0, \beta_1, \phi) = \sum_{i=1}^n \Bigg[ \log \left( \Gamma(y_i + \frac{1}{\phi}) \right) 
- \log \left( \Gamma(\frac{1}{\phi}) \right) - \log(y_i!) \\
+ \frac{1}{\phi} \log \left( \frac{1}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right) 
+ y_i \log \left( \frac{\exp(\beta_0 + \beta_1 x_i) \phi}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right) 
\Bigg]
$$



To find the estimating equation for \(\beta_1\), we differentiate this log-likelihood with respect to \(\beta_1\):

$$
\ell(\beta_1) = \sum_{i=1}^n \left[ \frac{1}{\phi} \log \left( \frac{1}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right) + y_i \log \left( \frac{\exp(\beta_0 + \beta_1 x_i) \phi}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right) \right]
$$

Differentiating both terms with respect to \(\beta_1\) yields:
$$
\frac{\partial \ell(\beta_1)}{\partial \beta_1} = \sum_{i=1}^n \left[ \frac{x_i \left( y_i - \exp(\beta_0 + \beta_1 x_i) \right)}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right]
$$

Setting this result equal to zero gives the estimating equation:

$$
\sum_{i=1}^n \left[ \frac{x_i \left( y_i - \exp(\beta_0 + \beta_1 x_i) \right)}{1 + \exp(\beta_0 + \beta_1 x_i) \phi} \right] = 0
$$




## Step 1: Negative Log-Likelihood Function

Since the `optim()` function in R minimizes functions, we use the negative log-likelihood function for the optimization process. The negative log-likelihood is simply the negative of the log-likelihood function.

```{r neg_log_likelihood,  echo=FALSE}
# Define the negative log-likelihood function for optimization
neg_log_likelihood <- function(params) {
  -log_likelihood(params, ants$abundance, ants$moisture)
}
```

## Step 2: Perform Optimization to Estimate Parameters

We now use the `optim()` function to find the maximum likelihood estimates (MLE) for \(\beta_0\), \(\beta_1\), and \(\phi\). The initial guesses for the parameters are provided, and the optimization is performed using the BFGS method.

```{r optimization, echo=FALSE, warning = FALSE}
# Define the negative log-likelihood function as a closure, referencing ants data directly
neg_log_likelihood <- function(params) {
  beta_0 <- params[1]  # The intercept parameter
  beta_1 <- params[2]  # The coefficient for moisture
  phi <- params[3]     # The overdispersion parameter
  
  # Calculate the mean (mu) for each observation
  mu <- exp(beta_0 + beta_1 * ants$moisture)
  
  # Compute the log-likelihood for each observation
  log_lik <- sum(
    lgamma(ants$abundance + 1 / phi) - lgamma(1 / phi) - lgamma(ants$abundance + 1) +
    (1 / phi) * log(1 / (1 + mu * phi)) +
    ants$abundance * log(mu * phi / (1 + mu * phi))
  )
  
  return(-log_lik)  # Return the negative log-likelihood for optimization
}

# Initial parameter guesses: beta_0 = 1, beta_1 = 0.1, phi = 0.5
initial_params <- c(1, 0.1, 0.5)

# Perform optimization to estimate beta_0, beta_1, and phi
optim_result <- optim(initial_params, neg_log_likelihood, method = "BFGS")

# Extract the estimated parameters
estimated_params <- optim_result$par

# Assign the estimates to beta_0_hat, beta_1_hat, and phi_hat
beta_0_hat <- estimated_params[1]
beta_1_hat <- estimated_params[2]
phi_hat <- estimated_params[3]

```

The estimated parameters provide insights into the relationship between soil moisture and ant abundance. The intercept (\(\beta_0\)) is `r beta_0_hat`, representing baseline abundance, while the coefficient for moisture (\(\beta_1\)) at `r beta_1_hat` shows the effect of soil moisture. The overdispersion parameter (\(\phi\)) at `r phi_hat` captures additional variability in the data.

# Log-Likelihood Optimization with Respect to \(\beta_1\)

In this question, we explore the relationship between \(\beta_1\) and the log-likelihood function, given the parameter estimates for \(\beta_0\) and \(\phi\). Our objective is to visualize how changes in \(\beta_1\) affect the log-likelihood, helping us understand its influence on the model.

```{r eval=TRUE, echo=FALSE}
# Given values
beta0 <- 2.509067
phi <- 2.289377
x_data <- ants$moisture  # Observed soil moisture data
y_data <- ants$abundance  # Observed ant abundance data

# Define the log-likelihood function in terms of beta1
log_likelihood <- function(beta1) {
  mu <- exp(beta0 + beta1 * x_data)
  sum(dnbinom(y_data, size = 1/phi, mu = mu, log = TRUE))
}

# Define a range of beta1 values
beta1_values <- seq(-5, 5, length.out = 100)

# Compute the log-likelihood for each beta1 value
log_likelihood_values <- sapply(beta1_values, log_likelihood)
```

In the Figure \@ref(fig:log-likelihood) above, we observe the log-likelihood values across a range of \(\beta_1\) values. The peak of this curve indicates the value of that maximizes the log-likelihood, corresponding to the most likely estimate of\(\beta_1\) given the data. This visualization aids in understanding how sensitive the model’s fit is to changes in \(\beta_1\).

```{r log-likelihood,echo=FALSE, fig.cap="Log likelihood as a function of beta 1", fig.pos="H", fig.width=5, fig.height=4}
# Plot the log-likelihood function
plot(beta1_values, log_likelihood_values, type = "l", 
     xlab = expression(beta[1]), 
     ylab = "Log-Likelihood", 
)

```

# Estimating the Optimal Value of \(\beta_1\)
To estimate the optimal value of \(\beta_1\) that maximizes the log-likelihood, we use the `optim()` function in R with the BFGS optimization method:

```{r, echo=FALSE} 
# Optimize to find beta1
result <- optim(par = 0, fn = log_likelihood, control = list(fnscale = -1), method = "BFGS")

# Extract the estimated beta1
beta1_estimate <- result$par
```
The estimated value of \(\beta_1\) is `r beta1_estimate`, which provides the most likely parameter value based on the observed data and model assumptions.

# Evaluating the Estimating Equation at \(\beta_1\)

To verify the estimated \(\beta_1\) value from Question 5, we compute the estimating equation at \(\beta_1\) to see if it approximates zero, as expected for a maximum likelihood estimate.

```{r, echo=FALSE}
# Define the estimating equation function
estimating_equation <- function(beta1, beta0, phi, x, y) {
  sum(x * (y - exp(beta0 + beta1 * x)) / (1 + exp(beta0 + beta1 * x) * phi))
}

# Calculate the value of the estimating equation at beta1_estimate
result <- estimating_equation(beta1_estimate, beta0, phi, x_data, y_data)
```

The resulting value of the estimating equation is **`r result`**. A result close to zero suggests that \(\beta_1\) is indeed an optimal estimate, indicating that the observed data align well with the expected values under the model.

# Visualizing the Mean-Variance Relationship

In this question, we use the estimated parameters to visualize the mean-variance relationship of predicted ant abundance and compare it to the observed data from Question 1.

## Step 1: Calculating Predicted Mean and Variance
Using the model’s estimated parameters, we calculate the predicted mean and variance of ant abundance for each soil moisture observation.

```{r, echo=FALSE}
# Estimated parameters from optimization
beta0 <- beta_0_hat
beta1 <- beta_1_hat
phi <- phi_hat

# Calculate the predicted mean (mu) and corresponding variance
predicted_mean <- exp(beta0 + beta1 * x_data) # Predicted mean for ant abundance
predicted_variance <- predicted_mean + (predicted_mean^2) * phi
```

## Step 2: Plotting the Predicted Mean-Variance Relationship
We plot the predicted mean against the predicted variance to visualize the model's mean-variance relationship for ant abundance.

## Step 3: Comparing with Observed Data
To evaluate the model fit, we overlay the actual mean-variance relationship from the observed data (calculated in Question 1) on the same plot. (Figure \@ref(fig:mean-var-relation2))

```{r mean-var-relation2, fig.cap="Mean-Variance Relationship of Predicted Ant Abundance", fig.pos="H", fig.width=5, fig.height=4,, eval=TRUE, echo=FALSE}
# Plot the Predicted Mean vs. Predicted Variance
plot(predicted_mean, predicted_variance, 
     xlab = "Predicted Mean of Ant Abundance",
     ylab = "Predicted Variance of Ant Abundance",
     pch = 16, col = "blue")
# Assuming mean_abundance and variance_abundance were computed in Question 1
points(mean_abundance, variance_abundance, pch = 1, col = "red")
legend("topright", legend = c("Predicted", "Actual"), pch = c(16, 1), col = c("blue", "red"))
```

## Interpretation
The plot shows the relationship between the mean and variance of the predicted abundance (in blue) and the actual observed data (in red). If the predicted values closely match the observed data, this suggests that the model’s mean-variance relationship aligns well with the data, supporting the negative binomial model’s appropriateness.